---
title: "Global Sea Ice EDA and Analysis"
author: "Jack Rabe"
date: "4/28/2022"
output: 
  html_document:
    toc: true
    toc_depth: 3
  
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(gt)
library(paletteer)
library(ggridges)
library(plotly)
library(sf)
library(ggmap)
library(tidycensus)
library(leaflet)
library(osmdata)
library(tigris)
library(ggsflabel)
library(ggthemes)
library(flexdashboard)
library(scales)
library(MetBrewer)
library(zoo)
library(rgdal)
library(raster)
library(rgeos)
library(rnaturalearth)
library(rnaturalearthdata)


#Working directory for .RMD
knitr::opts_knit$set(echo = TRUE,
                     root.dir = rprojroot::find_rstudio_root_file())

#Controlling figure output in markdown
knitr::opts_chunk$set(
#  fig.height =   
  fig.width = 6,
#  fig.asp = .5,
  out.width = "90%",
#  out.height = 
 fig.align  = "center",
  cache = FALSE,
  eval  = TRUE,
  echo  = TRUE,
  warning = FALSE
)

#My Colours (from viridis)
my_purple <- "#440154FF"
my_yellow <- "#FDE725FF"

#Set Theme for ggplot2
theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom"))

#Set Scientific notation output and decimal places for knitr
options(scipen = 999)
options(digits = 4)
options(dplyr.summarise.inform = FALSE)
```


Load in data

```{r}
#load in global data set
globe_ice.df <- read.csv(file = "./global/data/global_sea_ice_1979-2022.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("global_ice" = "Ice",
         "global_anomaly" = "Anomaly")

summary(globe_ice.df)

#load in northern hemisphere data set
nohem_ice.df <- read.csv(file = "./global/data/nohem_sea_ice_1979-2022.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("nohem_ice" = "Value",
         "nohem_anomaly" = "Anomaly")

summary(nohem_ice.df)

#load in southern hemisphere data set
sohem_ice.df <- read.csv(file = "./global/data/sohem_sea_ice_1979-2022.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("sohem_ice" = "Value",
         "sohem_anomaly" = "Anomaly")

summary(sohem_ice.df)

#load in hemispheric daily sea ice extent for duration calculations/plots
nohem_ice_daily.df <- read.csv(file = "./global/data/N_seaice_extent_daily_v3.0.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("nohem_extent" = "Extent") %>% 
  mutate(Date = as.Date(Date)) %>% 
  filter(!Date < "1988-12-31" & Year != 2022)  #remove up to last day of 1987, years prior sampled every other day, need last day for lag calculation, also remove 2022 since incomplete data
  

summary(nohem_ice_daily.df)

sohem_ice_daily.df <- read.csv(file = "./global/data/S_seaice_extent_daily_v3.0.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("sohem_extent" = "Extent") %>% 
  mutate(Date = as.Date(Date)) %>% 
  filter(!Date < "1988-12-31") %>%  #remove up to last day of 1987, years prior sampled every other day, need last day for lag calculation
  filter(!Date < "1988-12-31" & Year != 2022)  #remove up to last day of 1987, years prior sampled every other day, need last day for lag calculation, also remove 2022 since incomplete data

summary(sohem_ice_daily.df)

#load in sea level data
sea_level.df <- read.csv(file = "./global/data/epa-sea-level.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("mean_sea_level" = "CSIRO.Adjusted.Sea.Level",
         "lowerCI_sea_level" = "Lower.Error.Bound",
         "upperCI_sea_level" = "Upper.Error.Bound")

summary(sea_level.df)

#load in temp data
global_temp.df <- read.csv(file = "./global/data/monthly_globaltemp.csv", header = TRUE) %>% 
  as_tibble() %>% 
  arrange(-row_number()) %>%  #reverse row order so earliest years come first, matches sea ice df
  mutate(Date = as.Date(Date) %>% 
           as.yearmon()) #change to month year format
```


```{r}
#join the three dataframes
total_ice.df <- inner_join(globe_ice.df, nohem_ice.df,
                           by = c("Date" = "Date",
                                  "Date2" = "Date2",
                                  "Year" = "Year",
                                  "Month" = "Month")) 

total_ice.df <- inner_join(total_ice.df, sohem_ice.df,
                           by = c("Date" = "Date",
                                  "Date2" = "Date2",
                                  "Year" = "Year",
                                  "Month" = "Month"))

#tidy up data
total_ice.df <- total_ice.df %>% 
  filter(global_ice != -9999) %>% #remove rows with -9999 (no value)
  dplyr::select(!Date) %>% 
  rename(Date = Date2) %>% 
  mutate(Date = str_replace_all(string = Date, 
                            pattern = "/", 
                            replacement = "-"))

total_ice.df <- total_ice.df %>% 
  mutate(Date = mdy(Date))
  

#check how summary looks now
summary(total_ice.df)
str(total_ice.df)
```

# Plotting temporal trends in global and hemispheric sea ice change

```{r}
total_ice_bar.df <- total_ice.df %>% 
  pivot_longer(cols = ends_with("ice"), #taking from wide to long
               names_to = "ice_type", #category = type of event 
               values_to = "extent") %>%  #measurement = time till event
  filter(ice_type != "global_ice") %>% 
  mutate(ice_type = as.factor(ice_type) %>% 
           fct_recode("Northern" = "nohem_ice",
                      "Southern" = "sohem_ice"),
         text_label    = str_c(ice_type, #add as text aesthetic
                               "\n", Year, 
                               "\nSea Ice Extent - ", extent, "million km^2"))
  
#make ggplotly  
sea.ice.gg <- total_ice_bar.df %>% 
  ggplot(aes(x = Date, 
             y = extent, 
             fill = ice_type)) +
  geom_bar(aes(text = text_label), 
           stat = "identity", 
           position = "stack", 
           size = 1.5) +
  scale_y_continuous(limits = c(0,25)) +
  scale_x_date( #The date equiv of scale_x_continuous
    date_breaks = "5 year", #Breaks
    date_minor_breaks = "1 year", #Little lines in between
    date_labels = "%b '%y" #Label for the date,
  ) +
  xlab("") + ylab("Sea Ice Extent (millions km^2)") +
  scale_fill_manual("Hemisphere", values = met.brewer("Kandinsky", direction = 1))
  
  

#ggplotly() + tooltip to control label
ggplotly(sea.ice.gg, tooltip = "text") #tooltip controls text hover/label
```


```{r}
total_ice_line.df <- total_ice.df %>% 
  pivot_longer(cols = ends_with("ice"), #taking from wide to long
               names_to = "ice_type", #category = type of event 
               values_to = "extent") %>%  #measurement = time till event
  mutate(ice_type = as.factor(ice_type) %>% 
           fct_recode("Global" = "global_ice",
                      "Northern" = "nohem_ice",
                      "Southern" = "sohem_ice"))
```


Now make in plotly

```{r}
#make plotly of sea ice
seaice.plotly <- plot_ly(total_ice_line.df, x = ~Date, y = ~extent, color = ~ice_type)
seaice.plotly <- seaice.plotly %>% add_lines()
seaice.plotly <- seaice.plotly %>% layout(
    xaxis = list(
      rangeselector = list(
        buttons = list(
          list(
            count = 6,
            label = "6 mo",
            step = "month",
            stepmode = "backward"),
          list(
            count = 1,
            label = "1 yr",
            step = "year",
            stepmode = "backward"),
          list(
          count = 5,
            label = "5 yr",
            step = "year",
            stepmode = "backward"),
          list(
          count = 10,
            label = "10 yr",
            step = "year",
            stepmode = "backward"),
          list(
            count = 1,
            label = "YTD",
            step = "year",
            stepmode = "todate"),
          list(step = "all"))),

      rangeslider = list(type = "date")),
    
    yaxis = list(title = "Sea Ice Extent (millions km^2)"))

seaice.plotly
```


Let's make some plots looking at annual, as opposed to monthly, shifts in sea ice

```{r}
total_ice_year.df <- total_ice_line.df %>% 
  group_by(Year, ice_type) %>% 
  summarize(
    mean_ice_extent = mean(extent)
  ) %>% 
  ungroup() %>% 
  filter(Year != 2022) #remove last year since incomplete, throws off annual mean

#now plot
seaice.year.plotly <- plot_ly(data = total_ice_year.df, x = ~Year, y = ~mean_ice_extent, color = ~ice_type)
seaice.year.plotly <- seaice.year.plotly %>% add_lines()
seaice.year.plotly <- seaice.year.plotly %>% layout(
  xaxis = list("Year"),
  yaxis = list(title = "Sea Ice Extent (millions km^2)"))

seaice.year.plotly

```


Not surprisingly, it looks like we have a declining trend in sea ice extent. Let's create some simple linear models to compare trends.

```{r}
#make global dataset
global_ice_year.df <- total_ice_year.df %>% 
  filter(ice_type %in% "Global")

#make northern dataset
nohem_ice_year.df <- total_ice_year.df %>% 
  filter(ice_type %in% "Northern")

#make southern dataset
sohem_ice_year.df <- total_ice_year.df %>% 
  filter(ice_type %in% "Southern")

#construct three simple linear models comparing year with ice extent to get annual mean trend
mod1 <- lm(mean_ice_extent ~ Year, data = global_ice_year.df)
mod2 <- lm(mean_ice_extent ~ Year, data = nohem_ice_year.df)
mod3 <- lm(mean_ice_extent ~ Year, data = sohem_ice_year.df)

#check out model summaries
summary(mod1)
summary(mod2)
summary(mod3)
```

## When will sea ice disappear from the earth globally and in each hemisphere?

Based on these simple linear models, if mean annual sea ice trends continued at the same rate as they have since 1979, we would predict that average sea ice extent will reach 0 in the northern hemisphere by the year 2210, and it will reach 0 globally and, thus, in the southern hemisphere, by the year 2468.

Let's calculate the duration of sea ice expansion and and recession for each year

```{r}
#join both daily data frames
bothhem_ice_daily.df <- inner_join(nohem_ice_daily.df, sohem_ice_daily.df,
                           by = c("Date" = "Date",
                                  "Year" = "Year",
                                  "Month" = "Month",
                                  "Day" = "Day"))

#for northern hemisphere
bothhem_ice_daily.df <- bothhem_ice_daily.df %>% 
  mutate(nohem_ice_diff = nohem_extent - lag(nohem_extent),
         sohem_ice_diff = sohem_extent - lag(sohem_extent),
         nohem_ice_trend = case_when(
           nohem_ice_diff > 0 ~ "increasing",
           nohem_ice_diff %in% 0 ~ "no change",
           nohem_ice_diff < 0 ~ "decreasing"),
         sohem_ice_trend = case_when(
           sohem_ice_diff > 0 ~ "increasing",
           sohem_ice_diff %in% 0 ~ "no change",
           sohem_ice_diff < 0 ~ "decreasing")
  ) %>% 
  na.omit()
  
  
summary(bothhem_ice_daily.df)

#now summarize number of increasing and decreasing days for each year
#northern hemisphere
nohem_ice_duration.df <- bothhem_ice_daily.df %>% 
  group_by(Year, nohem_ice_trend) %>% 
  summarise(
    no_days = length(nohem_ice_trend)
  ) %>% 
  mutate(nohem_ice_trend = as.factor(nohem_ice_trend)) %>% 
  ungroup()


#southern hemisphere
sohem_ice_duration.df <- bothhem_ice_daily.df %>% 
  group_by(Year, sohem_ice_trend) %>% 
  summarise(
    no_days = length(sohem_ice_trend) 
  ) %>% 
  mutate(sohem_ice_trend = as.factor(sohem_ice_trend)) %>% 
  ungroup()

```


Now make plots showing trends in days if increasing and decreasing ice

```{r}
#northern hemisphere plot
nohem.ice.dur.plotly <- plot_ly(data = nohem_ice_duration.df, x = ~Year, y = ~no_days, color = ~nohem_ice_trend)
nohem.ice.dur.plotly <- nohem.ice.dur.plotly %>% add_lines()
nohem.ice.dur.plotly <- nohem.ice.dur.plotly %>% layout(
  xaxis = list("Year"),
  yaxis = list(title = "No. of Days"))

nohem.ice.dur.plotly

#southern hemisphere plot
sohem.ice.dur.plotly <- plot_ly(data = sohem_ice_duration.df, x = ~Year, y = ~no_days, color = ~sohem_ice_trend)
sohem.ice.dur.plotly <- sohem.ice.dur.plotly %>% add_lines()
sohem.ice.dur.plotly <- sohem.ice.dur.plotly %>% layout(
  xaxis = list("Year"),
  yaxis = list(title = "No. of Days"))

sohem.ice.dur.plotly
```


Now do make plots looking at sea ice anomaly data

```{r}
#alter df for anomaly plot
total_ice_anomaly.df <- total_ice.df %>% 
  pivot_longer(cols = ends_with("anomaly"), #taking from wide to long
               names_to = "ice_type", #category = type of event 
               values_to = "anomaly") %>%  #measurement = time till event
  mutate(ice_type = as.factor(ice_type) %>% 
           fct_recode("Global" = "global_anomaly",
                      "Northern" = "nohem_anomaly",
                      "Southern" = "sohem_anomaly"))

#make plotly of sea ice anomaly values
iceanomaly.plotly <- plot_ly(total_ice_anomaly.df, x = ~Date, y = ~anomaly, color = ~ice_type)
iceanomaly.plotly <- iceanomaly.plotly %>% add_lines()
iceanomaly.plotly <- iceanomaly.plotly %>% layout(
    xaxis = list(
      rangeselector = list(
        buttons = list(
          list(
            count = 6,
            label = "6 mo",
            step = "month",
            stepmode = "backward"),
          list(
            count = 1,
            label = "1 yr",
            step = "year",
            stepmode = "backward"),
          list(
          count = 5,
            label = "5 yr",
            step = "year",
            stepmode = "backward"),
          list(
          count = 10,
            label = "10 yr",
            step = "year",
            stepmode = "backward"),
          list(
            count = 1,
            label = "YTD",
            step = "year",
            stepmode = "todate"),
          list(step = "all"))),

      rangeslider = list(type = "date")),
    
    yaxis = list(title = "Sea Ice Anomalies"))

iceanomaly.plotly
```

It looks like global anomalies in sea ice are mostly driven by the northern hemisphere sea ice as opposed to the southern hemisphere. Let's rtun some simple linear models to check it out.


```{r}
#global anomaly vs southern hemisphere
mod4 <- lm(global_anomaly ~ sohem_anomaly, data = total_ice.df)
summary(mod4)

#global anomaly vs northern hemisphere
mod5 <- lm(global_anomaly ~ nohem_anomaly, data = total_ice.df)
summary(mod5)

#plots of both models above
#sohem
total_ice.df %>% 
  ggplot(aes(x = sohem_anomaly,  y = global_anomaly)) +
  stat_smooth(formula = "y ~ x",
              method = "lm")

#nohem
total_ice.df %>% 
  ggplot(aes(x = nohem_anomaly,  y = global_anomaly)) +
  stat_smooth(formula = "y ~ x",
              method = "lm")
```

As expected from an earlier visual assessment of anomaly data, the model with northern hemisphere anomaly explains 64.3% of variation in global anomaly data, relative to only 28% explained by the southern hemisphere anomalies.

# Modeling relationship between sea ice change and sea level rise

```{r}
#prep sea ice df for comparison with sea level data
total_ice_march.df <- total_ice.df %>% 
  filter(Month == 3) %>% #only use ice data from March, matches collectiion of annual sea level data
  dplyr::select(Year, global_ice, global_anomaly) %>% 
  mutate(global_ice_change = global_ice - lag(global_ice)) %>% #create annual change variable
  filter(Year != 1979) #remove first year with missing change value

#remove Date column
sea_level.df <- sea_level.df %>% 
  dplyr::select(!Date)

#join dataframes
sea_levelandice.df <- inner_join(total_ice_march.df, sea_level.df,
                                 by = ("Year" = "Year"))

#check out summary
summary(sea_levelandice.df)

#manual colors for legend
colors_icelevel <- c("Global Sea Level" = "#3399FF", "Global Sea Ice" = "#993399")

#plot sea ice and level together
sea_levelandice.df %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = global_ice, color = "Global Sea Ice")) +
  geom_line(aes(y = mean_sea_level*3, color = "Global Sea Level")) +
  geom_ribbon(aes(ymin = lowerCI_sea_level*3, ymax = upperCI_sea_level*3), fill = "#99CCFF", alpha = 0.25) +
  #add themes to clean up graph
  scale_y_continuous(
    # Features of the first axis
    name = "Global Sea Ice Extent (millions km^2)",
    # Add a second axis and specify its features
    sec.axis = sec_axis(trans = ~./3, name = "Mean Global Sea Level Change (in)")) +
  scale_color_manual("", #manual legend
                     values = colors_icelevel) + #set of colors matched with plot type
  theme(legend.position = "bottom",
        legend.key = element_rect(fill = "white"))

```



```{r}
#construct simple linear regression models
mod6 <- lm(mean_sea_level ~ global_ice, data = sea_levelandice.df)
mod7 <- lm(mean_sea_level ~ global_ice_change + poly(global_ice_change, 2), data = sea_levelandice.df)

#model summary
summary(mod6)
summary(mod7)

#check residuals vs fitted
par(mfrow = c(2,2))
plot(mod6)

par(mfrow = c(2,2))
plot(mod7)

#plot relationships
sea_levelandice.df %>% 
  ggplot(aes(x = global_ice,  y = mean_sea_level)) +
  stat_smooth(formula = "y ~ poly(x,2)",
              method = "lm")

sea_levelandice.df %>% 
  ggplot(aes(x = global_ice_change,  y = mean_sea_level)) +
  stat_smooth(formula = "y ~ poly(x,2)",
              method = "lm")
```


# Monthly sea ice and temperature

```{r}
#convert date to month year to match temperature data
total_ice_yearmon.df <- total_ice.df %>% 
  mutate(Date = as.yearmon(Date))

#split  temp df into two dfsbased on source
global_temp_gis.df <- global_temp.df %>% 
  filter(Source %in% "GISTEMP")

global_temp_gcag.df <- global_temp.df %>% 
  filter(Source %in% "GCAG")

#join dataframes
ice_temp_gis.df <- inner_join(total_ice_yearmon.df, global_temp_gis.df,
                          by = ("Date" = "Date")) 

ice_temp_gcag.df <- inner_join(total_ice_yearmon.df, global_temp_gcag.df,
                          by = ("Date" = "Date"))

#manual colors for legend
colors_icetemp <- c("Global Temperature" = "#3399FF", "Global Sea Ice" = "#993399")

#plot sea ice and level together
ice_temp_gis.df %>% 
  ggplot(aes(x = Date)) +
  geom_line(aes(y = global_ice, color = "Global Sea Ice")) +
  geom_line(aes(y = Mean*25, color = "Global Temperature")) +
  #add themes to clean up graph
  scale_y_continuous(
    # Features of the first axis
    name = "Global Sea Ice Extent (millions km^2)",
    # Add a second axis and specify its features
    sec.axis = sec_axis(trans = ~./25, name = "Mean Global Temp Change (degrees Celsius)")) +
  scale_color_manual("", #manual legend
                     values = colors_icetemp) + #set of colors matched with plot type
  theme(legend.position = "bottom",
        legend.key = element_rect(fill = "white"))

 #plot sea ice and level together
ice_temp_gis.df %>% 
  ggplot(aes(x = Date)) +
  geom_line(aes(y = global_anomaly, color = "Global Sea Ice")) +
  geom_line(aes(y = Mean*3, color = "Global Temperature")) +
  #add themes to clean up graph
  scale_y_continuous(
    # Features of the first axis
    name = "Global Sea Ice Anomalies",
    # Add a second axis and specify its features
    sec.axis = sec_axis(trans = ~./3, name = "Mean Global Temp Change (degrees Celsius)")) +
  scale_color_manual("", #manual legend
                     values = colors_icetemp) + #set of colors matched with plot type
  theme(legend.position = "bottom",
        legend.key = element_rect(fill = "white"))                   
```


```{r}
#construct simple linear regression models
mod8 <- lm(global_ice ~ Mean, data = ice_temp_gis.df)
mod9 <- lm(global_anomaly ~ Mean, data = ice_temp_gis.df)

#model summary
summary(mod8)
summary(mod9)

#check residuals vs fitted
par(mfrow = c(2,2))
plot(mod8)

par(mfrow = c(2,2))
plot(mod9)

#plot relationships
ice_temp_gis.df %>% 
  ggplot(aes(x = Mean,  y = global_ice)) +
  stat_smooth(formula = "y ~ x",
              method = "lm")

ice_temp_gis.df %>% 
  ggplot(aes(x = Mean,  y = global_anomaly)) +
  stat_smooth(formula = "y ~ x", 
              method = "lm")
```


# Spatial plots of sea ice

```{r, include=FALSE}
#Extract every data file about covid
ice_file_list <- list.files(path = "./global/data/Monthly Sea Ice Extent/all_spatial_files/", 
                            pattern = "[.]shp$", full.names = TRUE)

#check that it worked using gt() table
tibble(
  file_list = ice_file_list,
) %>% gt()
```


Read in a few csvs to get started

```{r}
#load shapefiles
seaice.jan1980 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_N_198001_polygon_v3.0.shp")
seaice.jan2020 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_N_202001_polygon_v3.0.shp")
seaice.sep1980 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_N_198009_polygon_v3.0.shp")
seaice.sep2020 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_N_202009_polygon_v3.0.shp")


# seaice.jan1980.transform <- stTransform(seaice.jan1980, CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))

#basemap
world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)
options(stringsAsFactors = FALSE)

world <- spTransform(world, "+proj=laea +lat_0=-90 +datum=WGS84")


#make basic plot of layers
ggplot() +
  geom_sf(data = seaice.jan1980, fill = "blue") +
  geom_sf(data = seaice.jan2020, fill = "purple")

ggplot() +
  geom_sf(data = seaice.sep1980, fill = "red") +
  geom_sf(data = seaice.sep2020, fill = "green") +
  geom_sf(data = world)

ggplot() +
  geom_sf(data = world) +
  geom_sf(data = seaice.jan2020, fill = "brown") +
  geom_sf(data = seaice.sep2020, fill = "yellow")
```


```{r}
#load shapefile
seaice.jan1980 <- readOGR("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_N_198001_polygon_v3.0.shp")

#check class and crs
class(seaice.jan1980)
crs(seaice.jan1980)
extent(seaice.jan1980)

seaice.jan1980.WGS84 <- spTransform(seaice.jan1980, CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))

crs(seaice.jan1980.WGS84)
extent(seaice.jan1980.WGS84)

broom::tidy(seaice.jan1980)
ggplot() +
  geom_polygon(data = seaice.jan1980, aes(x = long, y = lat), fill = "blue")
```

Make basemap

```{r}
#manual bounding box
seaice.box <- matrix(c(-180, 180, 38.01, 83.67),
                     ncol  = 2,
                     byrow = TRUE)
colnames(seaice.box) <- c("min", "max")
rownames(seaice.box) <- c("x", "y")
seaice.box

#Get the base map (foundational layer)
seaice_base.map <- get_map(
                       location = seaice.box,
                       source   = "stamen",
                       maptype  = "watercolor",
                       crop = TRUE
                      )
```


```{r}
ggmap(seaice_base.map) +
  geom_polygon(data = seaice.jan1980.WGS84, aes(x = long, y = lat)) +
  coord_map(projection = "orthographic")

leaflet(data = seaice.jan1980.WGS84) %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  addPolygons()
```











