---
title: "Flex dashboard"
author: ""
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
---

<style type="text/css"> .sidebar { overflow: auto; } </style>

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>
```{r setup, include = FALSE}
library(tidyverse)
library(lubridate)
library(gt)
library(paletteer)
library(ggridges)
library(plotly)
library(sf)
library(ggmap)
library(tidycensus)
library(leaflet)
library(osmdata)
library(tigris)
library(ggsflabel)
library(ggthemes)
library(flexdashboard)
library(scales)
library(MetBrewer)
library(zoo)
library(rgdal)
library(raster)
library(rgeos)
options(stringsAsFactors = FALSE)


#Working directory for .RMD
knitr::opts_knit$set(echo = TRUE,
                     root.dir = rprojroot::find_rstudio_root_file())

#Controlling figure output in markdown
knitr::opts_chunk$set(
#  fig.height =   
  fig.width = 6,
#  fig.asp = .5,
  out.width = "90%",
#  out.height = 
 fig.align  = "center",
  cache = FALSE,
  eval  = TRUE,
  echo  = TRUE,
  warning = FALSE
)

#My Colours (from viridis)
my_purple <- "#440154FF"
my_yellow <- "#FDE725FF"

#Set Theme for ggplot2
theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom"))

#Set Scientific notation output and decimal places for knitr
options(scipen = 999)
options(digits = 4)
options(dplyr.summarise.inform = FALSE)
```


```{r, loading data, include=FALSE}
#load in global data set
globe_ice.df <- read.csv(file = "./global/data/global_sea_ice_1979-2022.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("global_ice" = "Ice",
         "global_anomaly" = "Anomaly")

summary(globe_ice.df)

#load in northern hemisphere data set
nohem_ice.df <- read.csv(file = "./global/data/nohem_sea_ice_1979-2022.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("nohem_ice" = "Value",
         "nohem_anomaly" = "Anomaly")

summary(nohem_ice.df)

#load in southern hemisphere data set
sohem_ice.df <- read.csv(file = "./global/data/sohem_sea_ice_1979-2022.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("sohem_ice" = "Value",
         "sohem_anomaly" = "Anomaly")

summary(sohem_ice.df)

#load in hemispheric daily sea ice extent for duration calculations/plots
nohem_ice_daily.df <- read.csv(file = "./global/data/N_seaice_extent_daily_v3.0.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("nohem_extent" = "Extent") %>% 
  mutate(Date = as.Date(Date)) %>% 
  filter(!Date < "1988-12-31" & Year != 2022)  #remove up to last day of 1987, years prior sampled every other day, need last day for lag calculation, also remove 2022 since incomplete data
  

summary(nohem_ice_daily.df)

sohem_ice_daily.df <- read.csv(file = "./global/data/S_seaice_extent_daily_v3.0.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("sohem_extent" = "Extent") %>% 
  mutate(Date = as.Date(Date)) %>% 
  filter(!Date < "1988-12-31") %>%  #remove up to last day of 1987, years prior sampled every other day, need last day for lag calculation
  filter(!Date < "1988-12-31" & Year != 2022)  #remove up to last day of 1987, years prior sampled every other day, need last day for lag calculation, also remove 2022 since incomplete data

summary(sohem_ice_daily.df)

#load in sea level data
sea_level.df <- read.csv(file = "./global/data/epa-sea-level.csv", header = TRUE) %>% 
  as_tibble() %>% 
  rename("mean_sea_level" = "CSIRO.Adjusted.Sea.Level",
         "lowerCI_sea_level" = "Lower.Error.Bound",
         "upperCI_sea_level" = "Upper.Error.Bound")

summary(sea_level.df)

#load in temp data
global_temp.df <- read.csv(file = "./global/data/monthly_globaltemp.csv", header = TRUE) %>% 
  as_tibble() %>% 
  arrange(-row_number()) %>%  #reverse row order so earliest years come first, matches sea ice df
  mutate(Date = as.Date(Date) %>% 
           as.yearmon()) #change to month year format
```


```{r, joining data, include=FALSE}
#join the three dataframes
total_ice.df <- inner_join(globe_ice.df, nohem_ice.df,
                           by = c("Date" = "Date",
                                  "Date2" = "Date2",
                                  "Year" = "Year",
                                  "Month" = "Month")) 

total_ice.df <- inner_join(total_ice.df, sohem_ice.df,
                           by = c("Date" = "Date",
                                  "Date2" = "Date2",
                                  "Year" = "Year",
                                  "Month" = "Month"))

#tidy up data
total_ice.df <- total_ice.df %>% 
  filter(global_ice != -9999) %>% #remove rows with -9999 (no value)
  dplyr::select(!Date) %>% 
  rename(Date = Date2) %>% 
  mutate(Date = str_replace_all(string = Date, 
                            pattern = "/", 
                            replacement = "-"))

total_ice.df <- total_ice.df %>% 
  mutate(Date = mdy(Date))
  

#check how summary looks now
summary(total_ice.df)
str(total_ice.df)
```


```{r, barplot, include=FALSE}
total_ice_bar.df <- total_ice.df %>% 
  pivot_longer(cols = ends_with("ice"), #taking from wide to long
               names_to = "ice_type", #category = type of event 
               values_to = "extent") %>%  #measurement = time till event
  filter(ice_type != "global_ice") %>% 
  mutate(ice_type = as.factor(ice_type) %>% 
           fct_recode("Northern" = "nohem_ice",
                      "Southern" = "sohem_ice"),
         text_label    = str_c(ice_type, #add as text aesthetic
                               "\n", Year, 
                               "\nSea Ice Extent - ", extent, "million km^2"))
  
#make ggplotly  
sea.ice.gg <- total_ice_bar.df %>% 
  ggplot(aes(x = Date, 
             y = extent, 
             fill = ice_type)) +
  geom_bar(aes(text = text_label), 
           stat = "identity", 
           position = "stack", 
           size = 1.5) +
  scale_y_continuous(limits = c(0,25)) +
  scale_x_date( #The date equiv of scale_x_continuous
    date_breaks = "5 year", #Breaks
    date_minor_breaks = "1 year", #Little lines in between
    date_labels = "%b '%y" #Label for the date,
  ) +
  xlab("") + ylab("Sea Ice Extent (millions km^2)") +
  scale_fill_manual("Hemisphere", values = met.brewer("Kandinsky", direction = 1))
  
  

#ggplotly() + tooltip to control label
ggplotly(sea.ice.gg, tooltip = "text") #tooltip controls text hover/label
```



```{r, seaicemonthly, include=FALSE}
#create tidy data for producing line plotly of sea ice trends
total_ice_line.df <- total_ice.df %>% 
  pivot_longer(cols = ends_with("ice"), #taking from wide to long
               names_to = "ice_type", #category = scope of ice measurement
               values_to = "extent") %>%  #measurement = extent
  mutate(ice_type = as.factor(ice_type) %>% 
           fct_recode("Global" = "global_ice",
                      "Northern" = "nohem_ice",
                      "Southern" = "sohem_ice"))

#now make plotly of sea ice
seaice.plotly <- plot_ly(total_ice_line.df, x = ~Date, y = ~extent, color = ~ice_type)
seaice.plotly <- seaice.plotly %>% add_lines()
seaice.plotly <- seaice.plotly %>% layout(
    xaxis = list(
      rangeselector = list(
        buttons = list(
          list(
            count = 6,
            label = "6 mo",
            step = "month",
            stepmode = "backward"),
          list(
            count = 1,
            label = "1 yr",
            step = "year",
            stepmode = "backward"),
          list(
          count = 5,
            label = "5 yr",
            step = "year",
            stepmode = "backward"),
          list(
          count = 10,
            label = "10 yr",
            step = "year",
            stepmode = "backward"),
          list(
            count = 1,
            label = "YTD",
            step = "year",
            stepmode = "todate"),
          list(step = "all"))),

      rangeslider = list(type = "date")),
    
    yaxis = list(title = "Sea Ice Extent (millions km^2)"))

seaice.plotly
```


```{r, annualseaice, include=FALSE}
total_ice_year.df <- total_ice_line.df %>% 
  group_by(Year, ice_type) %>% 
  summarize(
    mean_ice_extent = mean(extent)
  ) %>% 
  ungroup() %>% 
  filter(Year != 2022) #remove last year since incomplete, throws off annual mean

#now plot
seaice.year.plotly <- plot_ly(data = total_ice_year.df, x = ~Year, y = ~mean_ice_extent, color = ~ice_type)
seaice.year.plotly <- seaice.year.plotly %>% add_lines()
seaice.year.plotly <- seaice.year.plotly %>% layout(
  xaxis = list("Year"),
  yaxis = list(title = "Sea Ice Extent (millions km^2)"))

seaice.year.plotly
```


```{r, anomalyplot, include=FALSE}
#alter df for anomaly plot
total_ice_anomaly.df <- total_ice.df %>% 
  pivot_longer(cols = ends_with("anomaly"), #taking from wide to long
               names_to = "ice_type", #category = type of event 
               values_to = "anomaly") %>%  #measurement = time till event
  mutate(ice_type = as.factor(ice_type) %>% 
           fct_recode("Global" = "global_anomaly",
                      "Northern" = "nohem_anomaly",
                      "Southern" = "sohem_anomaly"))

#make plotly of sea ice anomaly values
iceanomaly.plotly <- plot_ly(total_ice_anomaly.df, x = ~Date, y = ~anomaly, color = ~ice_type)
iceanomaly.plotly <- iceanomaly.plotly %>% add_lines()
iceanomaly.plotly <- iceanomaly.plotly %>% layout(
    xaxis = list(
      rangeselector = list(
        buttons = list(
          list(
            count = 6,
            label = "6 mo",
            step = "month",
            stepmode = "backward"),
          list(
            count = 1,
            label = "1 yr",
            step = "year",
            stepmode = "backward"),
          list(
          count = 5,
            label = "5 yr",
            step = "year",
            stepmode = "backward"),
          list(
          count = 10,
            label = "10 yr",
            step = "year",
            stepmode = "backward"),
          list(
            count = 1,
            label = "YTD",
            step = "year",
            stepmode = "todate"),
          list(step = "all"))),

      rangeslider = list(type = "date")),
    
    yaxis = list(title = "Sea Ice Anomalies"))

iceanomaly.plotly
```


```{r, include=FALSE}
#join both daily data frames
bothhem_ice_daily.df <- inner_join(nohem_ice_daily.df, sohem_ice_daily.df,
                           by = c("Date" = "Date",
                                  "Year" = "Year",
                                  "Month" = "Month",
                                  "Day" = "Day"))

#for northern hemisphere
bothhem_ice_daily.df <- bothhem_ice_daily.df %>% 
  mutate(nohem_ice_diff = nohem_extent - lag(nohem_extent),
         sohem_ice_diff = sohem_extent - lag(sohem_extent),
         nohem_ice_trend = case_when(
           nohem_ice_diff > 0 ~ "increasing",
           nohem_ice_diff %in% 0 ~ "no change",
           nohem_ice_diff < 0 ~ "decreasing"),
         sohem_ice_trend = case_when(
           sohem_ice_diff > 0 ~ "increasing",
           sohem_ice_diff %in% 0 ~ "no change",
           sohem_ice_diff < 0 ~ "decreasing")
  ) %>% 
  na.omit()
  
  
summary(bothhem_ice_daily.df)

#now summarize number of increasing and decreasing days for each year
#northern hemisphere
nohem_ice_duration.df <- bothhem_ice_daily.df %>% 
  group_by(Year, nohem_ice_trend) %>% 
  summarise(
    no_days = length(nohem_ice_trend)
  ) %>% 
  mutate(nohem_ice_trend = as.factor(nohem_ice_trend)) %>% 
  ungroup()


#southern hemisphere
sohem_ice_duration.df <- bothhem_ice_daily.df %>% 
  group_by(Year, sohem_ice_trend) %>% 
  summarise(
    no_days = length(sohem_ice_trend) 
  ) %>% 
  mutate(sohem_ice_trend = as.factor(sohem_ice_trend)) %>% 
  ungroup()
```


```{r, include=FALSE}
#northern hemisphere plot
nohem.ice.dur.plotly <- plot_ly(data = nohem_ice_duration.df, x = ~Year, y = ~no_days, color = ~nohem_ice_trend)
nohem.ice.dur.plotly <- nohem.ice.dur.plotly %>% add_lines()
nohem.ice.dur.plotly <- nohem.ice.dur.plotly %>% layout(
  xaxis = list("Year"),
  yaxis = list(title = "No. of Days"))

nohem.ice.dur.plotly

#southern hemisphere plot
sohem.ice.dur.plotly <- plot_ly(data = sohem_ice_duration.df, x = ~Year, y = ~no_days, color = ~sohem_ice_trend)
sohem.ice.dur.plotly <- sohem.ice.dur.plotly %>% add_lines()
sohem.ice.dur.plotly <- sohem.ice.dur.plotly %>% layout(
  xaxis = list("Year"),
  yaxis = list(title = "No. of Days"))

sohem.ice.dur.plotly
```


```{r, ice and sea level, include=FALSE}
#prep sea ice df for comparison with sea level data
total_ice_march.df <- total_ice.df %>% 
  filter(Month == 3) %>% #only use ice data from March, matches collectiion of annual sea level data
  dplyr::select(Year, global_ice, global_anomaly) %>% 
  mutate(global_ice_change = global_ice - lag(global_ice)) %>% #create annual change variable
  filter(Year != 1979) #remove first year with missing change value

#remove Date column
sea_level.df <- sea_level.df %>% 
  dplyr::select(!Date)

#join dataframes
sea_levelandice.df <- inner_join(total_ice_march.df, sea_level.df,
                                 by = ("Year" = "Year"))

#check out summary
summary(sea_levelandice.df)

#manual colors for legend
colors_icelevel <- c("Global Sea Level" = "#3399FF", "Global Sea Ice" = "#993399")

#plot sea ice and level together
iceandlevel.gg <- sea_levelandice.df %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = global_ice, color = "Global Sea Ice")) +
  geom_line(aes(y = mean_sea_level*3, color = "Global Sea Level")) +
  geom_ribbon(aes(ymin = lowerCI_sea_level*3, ymax = upperCI_sea_level*3), fill = "#99CCFF", alpha = 0.25) +
  #add themes to clean up graph
  scale_y_continuous(
    # Features of the first axis
    name = "Global Sea Ice Extent (millions km^2)",
    # Add a second axis and specify its features
    sec.axis = sec_axis(trans = ~./3, name = "Mean Global Sea Level Change (in)")) +
  scale_color_manual("", #manual legend
                     values = colors_icelevel) + #set of colors matched with plot type
  theme(legend.position = "bottom",
        legend.key = element_rect(fill = "white"))
```


```{r, ice and temp, include=FALSE}
#convert date to month year to match temperature data
total_ice_yearmon.df <- total_ice.df %>% 
  mutate(Date = as.yearmon(Date))

#split  temp df into two dfsbased on source
global_temp_gis.df <- global_temp.df %>% 
  filter(Source %in% "GISTEMP")

global_temp_gcag.df <- global_temp.df %>% 
  filter(Source %in% "GCAG")

#join dataframes
ice_temp_gis.df <- inner_join(total_ice_yearmon.df, global_temp_gis.df,
                          by = ("Date" = "Date")) 

ice_temp_gcag.df <- inner_join(total_ice_yearmon.df, global_temp_gcag.df,
                          by = ("Date" = "Date"))

#manual colors for legend
colors_icetemp <- c("Global Temperature" = "#3399FF", "Global Sea Ice" = "#993399")

#plot sea ice and level together
ice_temp.gg <- ice_temp_gis.df %>% 
  ggplot(aes(x = Date)) +
  geom_line(aes(y = global_ice, color = "Global Sea Ice")) +
  geom_line(aes(y = Mean*25, color = "Global Temperature")) +
  #add themes to clean up graph
  scale_y_continuous(
    # Features of the first axis
    name = "Global Sea Ice Extent (millions km^2)",
    # Add a second axis and specify its features
    sec.axis = sec_axis(trans = ~./25, name = "Mean Global Temp Change (degrees Celsius)")) +
  scale_color_manual("", #manual legend
                     values = colors_icetemp) + #set of colors matched with plot type
  theme(legend.position = "bottom",
        legend.key = element_rect(fill = "white"))

 #plot sea ice and level together
iceanom_temp.gg <- ice_temp_gis.df %>% 
  ggplot(aes(x = Date)) +
  geom_line(aes(y = global_anomaly, color = "Global Sea Ice")) +
  geom_line(aes(y = Mean*3, color = "Global Temperature")) +
  #add themes to clean up graph
  scale_y_continuous(
    # Features of the first axis
    name = "Global Sea Ice Anomalies",
    # Add a second axis and specify its features
    sec.axis = sec_axis(trans = ~./3, name = "Mean Global Temp Change (degrees Celsius)")) +
  scale_color_manual("", #manual legend
                     values = colors_icetemp) + #set of colors matched with plot type
  theme(legend.position = "bottom",
        legend.key = element_rect(fill = "white")) 
```


```{r, spatial sea ice, include=FALSE}
#load shapefiles
#northern hemisphere
Nseaice.mar1980 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_N_198003_polygon_v3.0.shp")
Nseaice.mar2020 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_N_202003_polygon_v3.0.shp")
Nseaice.sep1980 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_N_198009_polygon_v3.0.shp")
Nseaice.sep2020 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_N_202009_polygon_v3.0.shp")

#southern hemisphere
Sseaice.mar1980 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_S_198003_polygon_v3.0.shp")
Sseaice.mar2020 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_S_202003_polygon_v3.0.shp")
Sseaice.sep1980 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_S_198009_polygon_v3.0.shp")
Sseaice.sep2020 <- st_read("./global/data/Monthly Sea Ice Extent/all_spatial_files/extent_S_202009_polygon_v3.0.shp")

#make basic plot of layers
#northern hemisphere
#difference in ice between mar 1980 and mar 2020
Nspatice.mar80mar20.gg <- ggplot() +
  geom_sf(data = Nseaice.mar1980, fill = "#3399FF") +
  geom_sf(data = Nseaice.mar2020, fill = "#9900FF")

#difference in ice between sept 1980 and sept 2020
Nspatice.sep80sep20.gg <- ggplot() +
  geom_sf(data = Nseaice.sep1980, fill = "#FF6600") +
  geom_sf(data = Nseaice.sep2020, fill = "#33CC33")

#difference in ice between mar 2020 and sept 2020
Nspatice.mar20sep20.gg <- ggplot() +
  geom_sf(data = Nseaice.mar2020, fill = "#996633") +
  geom_sf(data = Nseaice.sep2020, fill = "#FFFF00")


#outhern hemisphere
#difference in ice between mar 1980 and mar 2020
Sspatice.mar80mar20.gg <- ggplot() +
  geom_sf(data = Sseaice.mar1980, fill = "#3399FF") +
  geom_sf(data = Sseaice.mar2020, fill = "#9900FF")

#difference in ice between sept 1980 and sept 2020
Sspatice.sep80sep20.gg <- ggplot() +
  geom_sf(data = Sseaice.sep1980, fill = "#FF6600") +
  geom_sf(data = Sseaice.sep2020, fill = "#33CC33")

#difference in ice between mar 2020 and sept 2020
Sspatice.mar20sep20.gg <- ggplot() +
  geom_sf(data = Sseaice.sep2020, fill = "#FFFF00") +
  geom_sf(data = Sseaice.mar2020, fill = "#996633")
```



Global Sea Ice Trends {style="position:relative;"}
========================================================

### <b><font face="Georgia" size="4em" color="#000000">About Sea Ice Data</font></b>
  The sea ice data you can find here is collected every day by the National Snow and Ice Data Center (NSIDC). Sea ice extent is divided into the northern and southern hemispheres, and a global measurement which is just the total of both hemispheres. It's measured as the total area (in millions of square kilometers) that sea ice covers. Anomaly values are based on how different an ice measurement is on a given date compared with the long-term average for that same day. Positve values indicate more ice than normal, and negative values indicate less ice than normal. The farther a value is from 0, the less normal it is.
  
### <b><font face="Georgia" size="6em" color="#000000">Monthly Sea Ice Extent</font></b>

```{r fig.width=10, fig.height=8, echo=FALSE}
seaice.plotly
```

### <b><font face="Georgia" size="4em" color="#000000">Monthly Sea Ice Extent Comments</font></b>
From looking at the plotly, we can see how the southern and northern hemisphere sea ice trends are about mirror images of each other. In other words, when northern sea ice is at its lowest, sea ice in the southern hemisphere tends to be close to its peak, and vice versa. However, we can see a slight lag between these opposite trends which can likely be attributed to the tilt of the Earth's axis. Interestingly, we also sea that sea ice has greater seasonal variation in the southern hemishpere than in the northern hemisphere.

### <b><font face="Georgia" size="6em" color="#000000">Monthly Sea Ice Anomalies</font></b>

```{r fig.width=10, fig.height=8, echo=FALSE}
iceanomaly.plotly
```

### <b><font face="Georgia" size="4em" color="#000000">Monthly Sea Ice Anomalies Comments</font></b>
From looking at the anomaly plot, it seems like global anomalies are mostly driven by changes in the northern hemisphere. Based on simple linear models, this rings true, as anomalies in the northern hemisphere explain ~64% of variation in global anomalies, while the southern hemisphere anomalies only explains around 28% of global variation> This makes sense when we look at the annual trends in sea ice extent, where global ice extent is mostly driven by declining sea ice in the northern hemisphere.

### <b><font face="Georgia" size="6em" color="#000000">Annual Trends in Sea Ice Extent</font></b>

```{r fig.width=10, fig.height=8, echo=FALSE}
seaice.year.plotly
```

### <b><font face="Georgia" size="4em" color="#000000">Annual Trends in Sea Ice Comments</font></b>
This plot shows the average sea ice extent for each year. Based on simple linear regression models, global sea ice decreased on average by ~49,000 square kilometers each year from 1979-2021. Sea ice on the northern hemisphere decreased on average by ~54,000 square kilometers. Based on these trends, if mean annual sea ice trends continued at the same rate as they have since 1979, we would predict that sea ice on average will disappear from the northern hemisphere by the year 2210, and globally / in the southern hemisphere, by the year 2468.

### <b><font face="Georgia" size="6em" color="#000000">Barplot of Hemishperic Sea Ice</font></b>

```{r fig.width=10, fig.height=8, echo=FALSE}
ggplotly(sea.ice.gg, tooltip = "text") #tooltip controls text hover/label
```

### <b><font face="Georgia" size="4em" color="#000000">Barplot of Hemishperic Sea Ice Comments</font></b>
This plot shows the same data as the first plot, but using stacked bars instead for easier monthly comparison of southern and northern hemisphere sea ice extent.


Sea Ice Duration  {style="position:relative;"}
========================================================

### <b><font face="Georgia" size="6em" color="#000000">Annual Duration of Expanding and Receding Sea Ice</font></b>
Surprisingly, unlike Minnesota lakes, the duration of sea ice expansion and recession in either the northern or southern hemisphere hasn't really changed over the last 30 years. However, even though durations haven't changed, we've shown from other plots that sea ice is expanding less and receding more when it does happen. Therefore, we suspect that the magnitude of increasing or decreasing is what's shifting, but not necessarily over the number of days in which those changes occur.

### <b><font face="Georgia" size="6em" color="#000000">Northern Hemisphere Sea Ice Durations</font></b>

```{r fig.width=10, fig.height=8, echo=FALSE}
nohem.ice.dur.plotly
```

### <b><font face="Georgia" size="6em" color="#000000">Southern Hemisphere Sea Ice Durations</font></b>

```{r fig.width=10, fig.height=8, echo=FALSE}
sohem.ice.dur.plotly
```


Sea Ice, Temperature, and Sea Level  {style="position:relative;"}
========================================================

### <b><font face="Georgia" size="6em" color="#000000">Relationship between Sea Ice and Temperature</font></b>
These two temperature plots show the significant effect that global mean temperature change has had on the extent of global sea icem and, even more, on the presence of anomalies in global sea ice. Essentially, this potentially demonstrates much of what we've already heard regarding the effects of global warming / climate change, that increasing global temperatures also increase extreme events and create more variation in trends like sea ice. 

### <b><font face="Georgia" size="6em" color="#000000">Global Sea Ice Extent and Temperature</font></b>

```{r fig.width=10, fig.height=8, echo=FALSE}
ice_temp.gg
```

### <b><font face="Georgia" size="6em" color="#000000">Global Sea Ice Anomalies and Temperature</font></b>

```{r fig.width=10, fig.height=8, echo=FALSE}
iceanom_temp.gg
```

### <b><font face="Georgia" size="6em" color="#000000">Global Sea Ice and Sea Level</font></b>

```{r fig.width=10, fig.height=8, echo=FALSE}
iceandlevel.gg
```

### <b><font face="Georgia" size="4em" color="#000000">Global Sea Ice and Sea Level Comments</font></b>
This plot shows very little association between global sea ice and temperature. However, we constructed a nonlinear model and found a small but significcant effect of sea ice on temperature, with sea level decreasingly slightly with increasing sea ice. It's important to note that because of data obtained, this plot represents sea ice and levels during the month of March.

Spatial Trends in Sea Ice {style="position:relative;"}
========================================================

This tab shows the variation in sea ice extent on both the northern (top panels) and southern hemisphere (bottom panels) over time. The first column shows how sea ice varies in March in each hemisphere from 1980 to 2020 to demostrate changes over the last 40 years. As we see, there are some losses in sea ice over time but it doesn't appear to be major at this time of year for either hemisphere. However, when we shift to the second column and look at ice trends from 1980 to 2020 during the month of September, we see a different story. 

In the northern hemisphere, sea ice looks to be cut in half over this forty year stretch, while the southern hemisphere still doesn't appear to decline very much, which explains the trends we saw in annual sea ice extent plot and linear model results, where the northern hemisphere sea ice was significantly declining but there was no significant decline in the southern hemisphere. 

Interestingly, when we shift to our final column, which looks at see ice differences between March 2020 and September 2020, we can see how variable sea ice is on a seasonal scale. This roughly depicts how sea ice changes at its maximum and minimum extent in each hemisphere. Of course, these exhibit inverse trends, with sea ice in the northern hemisphere reaching its greatest extent in March, and then its lowest in September, while in the southern hemisphere, sea ice roughly reaches its greatest extent in September and its lowest in March. This column shows that see ice in both hemispheres declines and increases dramatically (> 50%) every six months.


Row
-------------------------------------

### <b><font face="Georgia" size="4em" color="#000000">Northern Sea Ice: March 1980 (blue) - March 2020 (purple)</font></b>

```{r fig.width=6, fig.height=8, echo=FALSE}
Nspatice.mar80mar20.gg
```

### <b><font face="Georgia" size="4em" color="#000000">Southern Sea Ice: March 1980 (blue) - March 2020 (purple)</font></b>

```{r fig.width=6, fig.height=8, echo=FALSE}
Sspatice.mar80mar20.gg
```

Row
-------------------------------------

### <b><font face="Georgia" size="4em" color="#000000">Northern Sea Ice: September 1980 (orange) - September 2020 (green)</font></b>

```{r fig.width=6, fig.height=8, echo=FALSE}
Nspatice.sep80sep20.gg 
```

### <b><font face="Georgia" size="4em" color="#000000">Southern Sea Ice: September 1980 (orange) - September 2020 (green)</font></b>

```{r fig.width=6, fig.height=8, echo=FALSE}
Sspatice.sep80sep20.gg
```

Row
-------------------------------------

### <b><font face="Georgia" size="4em" color="#000000">Northern Sea Ice: March 2020 (brown) - September 2020 (yellow)</font></b>

```{r fig.width=6, fig.height=8, echo=FALSE}
Nspatice.mar20sep20.gg 
```

### <b><font face="Georgia" size="4em" color="#000000">Southern Sea Ice: March 2020 (brown) - September 2020 (yellow)</font></b>

```{r fig.width=6, fig.height=8, echo=FALSE}
Sspatice.mar20sep20.gg
```

